Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)\\ Technical Appendix Including Proofs and Details

### 2017-08-21 fully-abstract-comp/technical-report.tex
work in progress - trying to figure out what we wanted with return pointers. At the moment I am trying to introduce a code and data return pointer (I think that is what we ended up with).

%%% Named registers
\newcommand{\pcreg}{\mathrm{pc}}
\newcommand{\rstk}{\mathrm{r}_\mathrm{stk}}
\newcommand{\rO}{\mathrm{r}_\mathrm{ret}}
\newcommand{\rret}{\rO}
\newcommand{\rdata}{\mathrm{r}_\mathrm{data}}

@@

\begin{document}
\section{The two capability machines}
\subsection{Domains}
Source language domains:
\[
  \begin{array}{rrcl}
   \addrbnf,\basebnf \in & \Addr & \defeq & \nats \\
    \sealbasebnf, \sigma \in & \Seal & \defeq & \nats \\
    &\Word & \defeq & \ints \uplus \Caps\\
    \permbnf \in& \Perm & \defbnf & \dots \\
    &\glbnf & \defbnf & \glob \mid \local \\
    & \aendbnf & \in & \Addr \uplus \{\infty \} \\
    &  \sealendbnf & \in & \Seal \uplus \{\infty \} \\
    &\scbnf &\defbnf & ((\permbnf,\glbnf),\basebnf,\aendbnf,\addrbnf) \mid \seal{\glbnf,\sealbasebnf,\sealendbnf}\\
    & & & {\sourcecolor{} \mid \stkptr{\permbnf,\basebnf,\aendbnf,\addrbnf}}\\     &\cbnf & \defbnf &  \scbnf \mid \sealed{\scbnf,\sigma}{\sourcecolor{} \; \mid \retptr}\\ 
    &\RegName & \defbnf &\pcreg \mid \rO \mid \rstk \mid \rdata \mid \dots \\
    &\Reg & \defeq & \RegName \fun \Word\\
    &\Mem & \defeq & \Addr \fun \Word \\
    &\MemSeg & \defeq & \Addr \parfun \Word \\
    {\sourcecolor{} \sstkframe \in} & {\sourcecolor{} \StkFrame} & {\sourcecolor{} \defeq} & {\sourcecolor{} \Caps \times \MemSeg \times \Addr}\\
    {\sourcecolor{} \sstk \in}& {\sourcecolor{} \Stack} & {\sourcecolor{} \defeq} & {\sourcecolor{} \StkFrame^*} \\
    \Phi \in & \ExecConf & \defeq & \Mem \times \Reg {\sourcecolor{} \; \times \; \Stack \times \MemSeg }\\
    &\Conf & \defeq & \ExecConf \uplus \{\failed\} \uplus (\{\halted\} \times \Mem)
  \end{array}
\]
The target language domains are all the non blue parts in the above. The source language domains are the black and blue parts in the above. Further
\begin{itemize}
\item $\glbnf$ defines domain $\Global$
\item $\scbnf$ defines domain $\SealableCaps$
\item $\cbnf$ defines domain $\Caps$
\end{itemize}

@@

The source machine is also a capability machine with memory capabilities, local capabilities and sealed capabilities. Unlike the target machine, the source machine is going to have a built in stack along with special stack and return capabilities. The syntax of the source machine language is as follows:
\[
  \begin{array}{rcl}
    n & \in & \nats \\
    \src{r} &\in &  \RegName \setminus \{\rO,\rstk\}\\
    \src{\rn} &\defbnf & \src{r} \mid n \\
    \src{i} & \defbnf &  \trg{i} \mid \scall{\src{r}}{\src{r}}
  \end{array}
\]
There is one syntactic difference between the source language and the target language, namely the target language has an extra instruction in the $\scall{}{}$ instruction.

@@

\[
  \noRetStkReg{\reg} \defeq \forall r \in \reg \setminus \{ \pcreg, \rstk, \rO, \rdata \} \ldotp \reg(r) \neq \stkptr(\_)
\]

\[
  \noRetStkMs{\ms} \defeq \defeq \forall \aaddr \in \dom(\ms) \ldotp \ms(\aaddr) \neq \stkptr(\_)
\]

@@

If we are not in one of the two cases above, then the execution fails.
\begin{align*}
  \ssem{\scall{r_1}{r_2}} = & 
    \begin{cases}
      (\mem,\reg',\stk',\ms_\stk') & \arraycolsep=0pt
                                      \begin{array}[t]{l}
                                        % "red conditions" (reference to blackboard picture)
                                        \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \\
                                        \Phi(r_1) = \sealed{\cb_1,\sigma} \text{ and } \\
                                        \Phi(r_2) = \sealed{\cb_2,\sigma} \text{ and } \\
                                        \nonExec{cb_2} \text{ and } \\
                                        \Phi(\rstk) = \stkptr{\perm, \stdrng, \aaddr} \text{ and } \\
                                        % "yellow conditions"
                                        \cb_2 \neq \stkptr{\_,\_,\_,\_} \text{ and } \\
                                        \noRetStkReg{\reg} \text{ and } \\
                                        \noRetStkMs{\ms_\stk'} \text{ and } \\
                                        \nonLoc{\cb_1,\cb_2} \text{ and } \\
                                        \reg' = \reg
                                        \begin{array}[t]{l}
                                        \update{\pcreg}{\cb_1}\update{\rstk}{\stkptr{\perm,[\aaddr,eaddr],\aaddr}}\\
                                        \update{\rO}{\retptr}\update{\rdata}{\cb_2} \text{ and }
                                        \end{array}
 \\
                                        \opc = \Phi(\pcreg)\update{a}{\Phi(\pcreg).a+\calllen} \text{ and } \\
                                        \ms_\priv = \ms_\stk \setminus [\aaddr,\eaddr] \text{ and } \\
                                        \stk' = (\opc,\ms_\priv,\perm,\stdrng,\aaddr) :: \stk \text{ and } \\
                                        \ms_\stk' = \ms_\stk |_{[\baddr,\eaddr]}
                                      \end{array}\\
                                      & \\
      \undefined & \arraycolsep=0pt
                                      \begin{array}[t]{l}
                                        % "red conditions" (reference to blackboard picture)
                                        \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \\
                                        \Phi(r_1) = \sealed{\cb_1,\sigma} \text{ and } \\
                                        \Phi(r_2) = \sealed{\cb_2,\sigma} \text{ and } \\
                                        \nonExec{cb_2} \text{ and } \\
                                        \Phi(\rstk) = \stkptr{\perm, \stdrng, \aaddr} \text{ and } \\
                                        % "yellow conditions"
                                        \text{not } \arraycolsep=0pt
                                        \left(\begin{array}{l}
                                          \cb_2 \neq \stkptr{\_,\_,\_,\_} \text{ and } \\
                                          \noRetStkReg{\reg} \text{ and } \\
                                          \noRetStkMs{\ms_\stk'} \text{ and } \\
                                          \nonLoc{\cb_1,\cb_2}
                                        \end{array} \right)
                                      \end{array}\\
                                      & \\
      \failed & \text{otherwise}
    \end{cases}
\end{align*}

>>>>

%%% Named registers
\newcommand{\pcreg}{\mathrm{pc}}
\newcommand{\rstk}{\mathrm{r}_\mathrm{stk}}
\newcommand{\rO}{\mathrm{r}_\mathrm{ret}}
\newcommand{\rret}{\rO}
\newcommand{\rretc}{\mathrm{r}_\mathrm{ret code}}
\newcommand{\rretd}{\mathrm{r}_\mathrm{ret data}}
\newcommand{\rdata}{\mathrm{r}_\mathrm{data}}

@@

\begin{document}
\section{The two capability machines}
\subsection{Domains}
Source language domains:
\[
  \begin{array}{rrcl}
   \addrbnf,\basebnf \in & \Addr & \defeq & \nats \\
    \sealbasebnf, \sigma \in & \Seal & \defeq & \nats \\
    &\Word & \defeq & \ints \uplus \Caps\\
    \permbnf \in& \Perm & \defbnf & \dots \\
    &\glbnf & \defbnf & \glob \mid \local \\
    & \aendbnf & \in & \Addr \uplus \{\infty \} \\
    &  \sealendbnf & \in & \Seal \uplus \{\infty \} \\
    &\scbnf &\defbnf & ((\permbnf,\glbnf),\basebnf,\aendbnf,\addrbnf) \mid \seal{\glbnf,\sealbasebnf,\sealendbnf}\\
    & & & {\sourcecolor{} \mid \stkptr{\permbnf,\basebnf,\aendbnf,\addrbnf}}\\ 
    &\cbnf & \defbnf &  \scbnf \mid \sealed{\scbnf,\sigma}{\sourcecolor{} \; \mid \retptrd \mid \retptrc}\\ 
    &\RegName & \defbnf &\pcreg \mid \rretc \mid \rretd \mid \rstk \mid \rdata \mid \dots \\
    &\Reg & \defeq & \RegName \fun \Word\\
    &\Mem & \defeq & \Addr \fun \Word \\
    &\MemSeg & \defeq & \Addr \parfun \Word \\
    {\sourcecolor{} \sstkframe \in} & {\sourcecolor{} \StkFrame} & {\sourcecolor{} \defeq} & {\sourcecolor{} \Caps \times \Caps \times \MemSeg \times \Addr}\\
    \src{\sstk \in}& \src{ \Stack} & \src{ \defeq} & \src{ \StkFrame^*} \\
    \Phi \in & \ExecConf & \defeq & \Mem \times \Reg {\sourcecolor{} \; \times \; \Stack \times \MemSeg }\\
    &\Conf & \defeq & \ExecConf \uplus \{\failed\} \uplus (\{\halted\} \times \Mem)
  \end{array}
\]
The target language domains are all the non blue parts in the above. The source language domains are the black and blue parts in the above. Further
\begin{itemize}
\item $\glbnf$ defines domain $\Global$
\item $\scbnf$ defines domain $\SealableCaps$
\item $\cbnf$ defines domain $\Caps$
\item $\Perm$ is defined as the permissions in Figure~\ref{fig:perm-hier}.
\end{itemize}

In the source language, $\src{\Stack}$ is a call stack that contains the data for each call. The call stack consists of a number of $\src{\StkFrame}$'s that contains 1) the old pc, 2) the old data capability, 3) caller private stack, and 4) the address of the old stack pointer.

@@

The source machine is also a capability machine with memory capabilities, local capabilities and sealed capabilities. Unlike the target machine, the source machine is going to have a built in stack along with special stack and return capabilities. The syntax of the source machine language is as follows:
\[
  \begin{array}{rcl}
    n & \in & \nats \\
    \src{r} &\in &  \RegName \setminus \{\rretc, \rretd, \rstk\}\\
    \src{\rn} &\defbnf & \src{r} \mid n \\
    \src{i} & \defbnf &  \trg{i} \mid \scall{\src{r}}{\src{r}}
  \end{array}
\]
There is one syntactic difference between the source language and the target language, namely the target language has an extra instruction in the $\scall{}{}$ instruction.

@@

\[
  \noRetStkReg{\reg} \defeq \forall r \in \reg \setminus \{ \pcreg, \rstk, \rretc, \rretd, \rdata \} \ldotp \reg(r) \neq \stkptr{\_}
\]

\[
  \noRetStkMs{\ms} \defeq \forall \aaddr \in \dom(\ms) \ldotp \ms(\aaddr) \neq \stkptr{\_}
\]

@@

If we are not in one of the two cases above, then the execution fails.
\begin{align*}
  \ssem{\scall{r_1}{r_2}} = & 
    \begin{cases}
      (\mem,\reg',\stk',\ms_\stk') & \arraycolsep=0pt
                                      \begin{array}[t]{l}
                                        % "red conditions" (reference to blackboard picture)
                                        \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \\
                                        \Phi(r_1) = \sealed{\cb_1,\sigma} \text{ and } \\
                                        \Phi(r_2) = \sealed{\cb_2,\sigma} \text{ and } \\
                                        \nonExec{cb_2} \text{ and } \\
                                        \Phi(\rstk) = \stkptr{\perm, \stdrng, \aaddr} \text{ and } \\
                                        % "yellow conditions"
                                        \cb_2 \neq \stkptr{\_,\_,\_,\_} \text{ and } \\
                                        \noRetStkReg{\reg} \text{ and } \\
                                        \noRetStkMs{\ms_\stk'} \text{ and } \\
                                        \nonLoc{\cb_1,\cb_2} \text{ and } \\
                                        \reg' = \reg
                                        \begin{array}[t]{l}
                                        \update{\pcreg}{\cb_1}\update{\rdata}{\cb_2}\\
                                        \update{\rretc}{\retptrc}\\
                                        \update{\rretd}{\retptrd}\\
                                        \update{\rstk}{\stkptr{\perm,[\aaddr,\eaddr],\aaddr}} \text{ and }
                                        \end{array}
 \\
                                        \opc = \Phi(\pcreg)\update{a}{\Phi(\pcreg).a+\calllen} \text{ and } \\
                                        \ms_\priv = \ms_\stk \setminus [\aaddr,\eaddr] \text{ and } \\
                                        \stk' = (\opc,\Phi(\rdata),\ms_\priv,\perm,\stdrng,\aaddr) :: \stk \text{ and } \\
                                        \ms_\stk' = \ms_\stk |_{[\baddr,\eaddr]}
                                      \end{array}\\
                                      & \\
      \undefined & \arraycolsep=0pt
                                      \begin{array}[t]{l}
                                        % "red conditions" (reference to blackboard picture)
                                        \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \\
                                        \Phi(r_1) = \sealed{\cb_1,\sigma} \text{ and } \\
                                        \Phi(r_2) = \sealed{\cb_2,\sigma} \text{ and } \\
                                        \nonExec{cb_2} \text{ and } \\
                                        \Phi(\rstk) = \stkptr{\perm, \stdrng, \aaddr} \text{ and } \\
                                        % "yellow conditions"
                                        \text{not } \arraycolsep=0pt
                                        \left(\begin{array}{l}
                                          \cb_2 \neq \stkptr{\_,\_,\_,\_} \text{ and } \\
                                          \noRetStkReg{\reg} \text{ and } \\
                                          \noRetStkMs{\ms_\stk'} \text{ and } \\
                                          \nonLoc{\cb_1,\cb_2}
                                        \end{array} \right)
                                      \end{array}\\
                                      & \\
      \failed & \text{otherwise}
    \end{cases}
\end{align*}
\lau{do we need to require anything on the data capability we put on the call stack?}
